---
output: github_document
bibliography: vignettes/References.bib
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd -->

[![CRAN status](https://www.r-pkg.org/badges/version/IEATools)](https://cran.r-project.org/package=IEATools)
[![lifecycle](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://www.tidyverse.org/lifecycle/#stable)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![R-CMD-check](https://github.com/MatthewHeun/IEATools/workflows/R-CMD-check/badge.svg)](https://github.com/MatthewHeun/IEATools/actions)
[![Codecov test coverage](https://codecov.io/gh/MatthewHeun/IEATools/branch/master/graph/badge.svg)](https://codecov.io/gh/MatthewHeun/IEATools?branch=master)
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.5086371.svg)](https://doi.org/10.5281/zenodo.5086371)


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```


# IEATools

The `R` package `IEATools` provides 
tools to interact with data from the [International Energy Agency (IEA)](http://www.iea.org). 


## Installation

<!-- You can install `IEATools` from CRAN with: -->

<!-- ```{r CRAN-installation, eval = FALSE} -->
<!-- install.packages("IEATools") -->
<!-- ``` -->

<!-- You can install a recent development version of `IEATools` from github with: -->
You can install `IEATools` from github with:

```{r gh-installation, eval = FALSE}
# install devtools if not already installed
# install.packages("devtools")
devtools::install_github("MatthewHeun/IEATools")
# To build vignettes locally, use
devtools::install_github("MatthewHeun/IEATools", build_vignettes = TRUE)
```


## History

The functions in this package were used in 
[Heun et al. [-@Heun:2018]](https://doi.org/10.1016/j.apenergy.2018.05.109).

Permission to distribute IEA extended energy balance data
for two years and two countries was obtained 
by Matthew Kuperus Heun from Nick Johnstone of the IEA
during a phone call at 11 AM on Mon, 3 June 2019.


## More Information

Find more information, including vignettes and function documentation, at 
<https://MatthewHeun.github.io/IEATools/>.




## References
