---
output: github_document
bibliography: vignettes/References.bib
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd -->

[![CRAN status](https://www.r-pkg.org/badges/version/IEATools)](https://cran.r-project.org/package=IEATools)
[![lifecycle](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://www.tidyverse.org/lifecycle/#stable)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![R-CMD-check](https://github.com/MatthewHeun/IEATools/workflows/R-CMD-check/badge.svg)](https://github.com/MatthewHeun/IEATools/actions)
[![Codecov test coverage](https://codecov.io/gh/MatthewHeun/IEATools/branch/master/graph/badge.svg)](https://codecov.io/gh/MatthewHeun/IEATools?branch=master)
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.5086371.svg)](https://doi.org/10.5281/zenodo.5086371)


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```


# IEATools


## Statement of need

Extended World Energy Balance (EWEB) data from the
[International Energy Agency (IEA)](http://www.iea.org) (IEA) 
are indispensable for societal energy analysis and
societal exergy analysis (SEA).
However, 

* the EWEB data are not available native in the `R` data frame format;
* as distributed, the EWEB data are definitely not 
  [tidy](http://vita.had.co.nz/papers/tidy-data.pdf); and 
* the EWEB are not in physical supply-use table (PSUT) format
  proposed by [@Heun:2018](https://doi.org/10.1016/j.apenergy.2018.05.109).

`IEATools` provides functions to address the above needs.


## Installation

<!-- You can install a recent development version of `IEATools` from github with: -->
You can install `IEATools` from GitHub with:

```{r gh-installation, eval = FALSE}
# install devtools if not already installed
# install.packages("devtools")
devtools::install_github("MatthewHeun/IEATools")
# To build vignettes locally, use
devtools::install_github("MatthewHeun/IEATools", build_vignettes = TRUE)
```


## History

The functions in this package were used in 
[Heun et al. [-@Heun:2018]](https://doi.org/10.1016/j.apenergy.2018.05.109).

Permission to distribute IEA extended energy balance data
for two years and two countries was obtained 
by Matthew Kuperus Heun from Nick Johnstone of the IEA
during a phone call at 11 AM on Mon, 3 June 2019.


## More Information

Find more information, including vignettes and function documentation, at 
<https://MatthewHeun.github.io/IEATools/>.




## References
