---
title: "Final to Useful Analysis"
author: "Matthew Kuperus Heun"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Final to Useful Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
library(IEATools)
```

<!-- Establish some helpful LaTeX shortcuts for equations -->
\newcommand{\transpose}[1]{#1^\mathrm{T}}
\newcommand{\inverse}[1]{#1^{\mathrm{-}1}}
\newcommand{\mat}[1]{\mathbf{#1}}
\newcommand{\colvec}[1]{\mathbf{#1}}
\newcommand{\rowvec}[1]{\transpose{\colvec{#1}}}
\newcommand{\inversetranspose}[1]{\transpose{\left( \inverse{\mat{#1}} \right)}}
\newcommand{\transposeinverse}[1]{\inverse{\left( \transpose{\mat{#1}} \right)}}
\newcommand{\hatinv}[1]{\inverse{\widehat{#1}}}


## Introduction

Extended energy balance data from the 
[International Energy Agency (IEA)](https://www.iea.org)
provides both primary and final energy information.
(Primary energy is extracted from the biosphere, and 
final energy is sold to consumers.)
But consumers don't buy final energy for its own sake.
They buy final energy so they can convert it to useful energy 
(heat, light, work) that, 
when combined with infrastructure 
(buildings, roads, etc.),
provides energy services
(heated and cooled space, passenger and freight transport, etc.).

So some energy conversion chain analyses extend the final energy provided by the IEA 
to useful energy through a process that involves

(a) allocating final energy to final-to-useful energy conversion machines,
(b) defining the efficiency of those final-to-useful energy conversion machines, and
(c) calculating the useful energy injected into the the final demand economic sectors.

The `IEATools` package provides several functions to assist the process of pushing 
from final energy to useful energy.
This vignette demonstrates their use
and suggests a workflow for this process.


## Allocation

Extending the IEA's extended energy balance data to the useful stage begins 
with defining allocations of final energy carriers to final-to-useful energy conversion machines.
To demonstrate this process, we'll use the sample IEA data bundled with this package.

```{r}
library(dplyr)
library(IEATools)
Tidy_iea_data <- load_tidy_iea_df() %>% 
  specify_all() %>% 
  glimpse()
```

The next step is to develop a blank template in which the analyst 
can provide final-to-useful allocation values (`C`) in percentages.
`fu_allocation_template()` builds a final-to-useful allocation template.

```{r}
FU_allocation_template <- load_tidy_iea_df() %>% 
  specify_all() %>%
  fu_allocation_template()
FU_allocation_template %>% 
  glimpse()
```

To obtain a nicely-formatted Excel spreadsheet, use the
`write_fu_allocation_template()` function.

```{r}
# FU_allocation_template %>%
#   write_fu_allocation_template("~/Desktop/Test_FU_allocation_template.xlsx")
```

The analyst should complete all of the unshaded cells in the `Machine` and `Eu.product` columns.
`Machine` should contain names of final-to-useful machines such as "Automobiles," "Industrial heat/furnace," etc.
`Eu.product` should be filled with names of useful energy carriers such as "HTH.600.C", "MD", or "Light".
Note that later processing will parse the heat categories for their temperature, and the required format is `*TH.nnn.U`, where

* `*` (the first character) is arbitrary, 
  but `H`, `M`, and `L` are typical for high, medium, and low temperature heat, respectively,
* `TH.` is required,
* `nnn` can be any number of digits, optionally preceded by a negative sign, 
* `.` is required, and
* `U` is the temperature unit, one of `C`, `F`, `R`, or `K` for degrees Celsius, degrees Fahrenheit, 
  rankine, or kelvin, respectively.

The `C` rows representation allocations of `E.dot` energy to final-to-useful machines for each year of available IEA data.
Three allocation rows (`C_1 [%]`, `C_2 [%]`, and `C_3 [%]`) are provided for each combination 
of `Flow.aggregation.point`, `Ef.product`, and `Destination`.
Not all `C_i [%]` rows need to be filled,
and the analyst can add more allocation rows (`C_4 [%]`, `C_5 [%]`, ...) as required.

The rows are in IEA order for `Flow.aggregation.point`, `Ef.product`, and `Destination`.

The `Maximum.values` column provides an indication of which `Destination`s receive more energy than others,
allowing the analyst to focus more attention on larger energy flows, as desired.


## Final-to-useful efficiencies and useful exergy-to-energy ratios

After completing the final-to-useful allocation table,
the analyst must specify the efficiencies of the final-to-useful machines.
To assist with this task, the `eta_fu_template()` function can be invoked.
To demonstrate, we'll use `load_fu_allocation_data()`, 
a function that reads a completed allocation table out of an excel file.
The default `path` argument of `load_fu_allocation_data()`
points to a completed final-to-useful allocation table
that is provided with `IEATools`.

```{r}
Eta_fu_template <- load_fu_allocation_data() %>% 
  eta_fu_template()
Eta_fu_template %>% 
  glimpse()
```

To obtain a nicely-formatted Excel spreadsheet, use the
`write_fu_allocation_template()` function.

```{r}
# Eta_fu_template %>%
#   write_eta_fu_template("~/Desktop/Test_eta_fu_template.xlsx")
```
The analyst should fill all unshaded cells in `eta.fu` rows with 
final-to-useful energy efficiency values (in %). 
`phi.u` rows should be completed with exergy-to-energy ratios for the useful stage energy carrier.
Note that many `phi.u` values are already filled for cases where the 
phi value is known or easily calculable, 
specifically mechanical drive (phi = 1) and heat (phi = 1 - T_0/T_H, where T_H is the heat temperature
embedded in the `*TH.nnn.U` useful exergy string).

The `Maximum.values` column provides an indication of which `Destination`s receive more energy than others,
allowing the analyst to focus more attention on larger energy flows, as desired.
The `E.dot_machine` and `E.dot_machine [%]` rows indicate energy flows and percentages
into each final-to-useful machine in each year.

After allocations and efficiencies are specified (by completing the templates),
functions can be used to move the last stage of an energy conversion chain 
from final energy to useful energy.


## Allocation (C) matrices

Next, the `C` values from the allocation table must be arranged into $\mat{C}$ matrices
using the `form_C_mats()` function.


```{r}
C_mats <- load_fu_allocation_data() %>% 
  form_C_mats()
glimpse(C_mats)
```

Note that two types of $\mat{C}$ matrices are formed, 
$\mat{C}_{EIOU}$ and $\mat{C}_Y$.
Both are needed later.


## Efficiency and exergy-to-energy ratio vectors

Next, vectors containing final-to-useful energy efficiencies ($\colvec{\eta}_{fu}$) and
exergy-to-energy ratios ($\colvec{\phi}$) must be formed
using the `form_eta_fu_phi_u_vecs()` function.

```{r}
eta_fu_vecs <- load_eta_fu_data() %>% 
  form_eta_fu_phi_u_vecs()
glimpse(eta_fu_vecs)
```


## Extend to useful

To extend an energy conversion chain (ECC) to the useful stage, 
first convert the IEA data into physical supply-use table format
with final energy being the last stage of the energy conversion chain.
Thereafter, join the $\mat{C}$ matrices and $\colvec{\eta_{fu}}$ vectors
to the data frame of PSUT matrices.

```{r}
# Get a list of the metadata columns.
# This list will be the same for C_mats.
m_cols <- eta_fu_vecs %>% 
  meta_cols(return_names = TRUE,
            years_to_keep = IEATools::iea_cols$year,
            not_meta = c(IEATools::template_cols$eta_fu, IEATools::template_cols$phi_u))
m_cols
psut_mats_final <- load_tidy_iea_df() %>% 
  specify_all() %>% 
  prep_psut() %>% 
  dplyr::full_join(C_mats, by = m_cols) %>% 
  dplyr::full_join(eta_fu_vecs, by = m_cols)
glimpse(psut_mats_final)
```

Then use the allocation and efficiency matrices to extend to the useful stage
with the `extend_to_useful()` function.

```{r}
psut_mats_useful <- psut_mats_final %>% 
  extend_to_useful()
glimpse(psut_mats_useful)
```

Notice that several additional rows have been created in the data frame.
The additional rows contain versions of ECC matrices wherein the last stage of the ECC
is useful energy.
Extending to the useful stage means many additional energy products
are included in the energy conversion chain. 
For example, the $\mat{Y}$ matrix includes additional rows for
mechanical drive (MD), heat (MTH.100.C, e.g.), and light.

```{r}
# A Y matrix where final energy is the last stage of the energy conversion chain
rownames(psut_mats_useful$Y[[1]])

# A Y matrix where useful energy is the last stage of the energy conversion chain
rownames(psut_mats_useful$Y[[5]])
```

<!-- It would be nice to calculate the energy balance of all rows of the data frame here,
     showing that extending to the useful stage does not affect the energy balance of the ECC
     when done correctly.  
     Unfortunately, the function to perform those energy balances resides in the Recca package. 
     By design, IEATools cannot depend on Recca. -->


## Summary

The process of moving from final energy to useful energy as the last stage of an energy conversion chain
is aided by several functions in the `IEATools` package.
In particular,

* `prep_psut()`,
* `form_C_mats()`,
* `form_eta_fu_phi_u_vecs()`, and
* `extend_to_useful()`

provide a convenient workflow for that process.