---
title: "Final to Useful Analysis"
author: "Matthew Kuperus Heun"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Final to Useful Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
library(IEATools)
```


## Introduction

Extended energy balance data from the 
[International Energy Agency (IEA)](https://www.iea.org)
provides both primary and final energy information.
(Primary energy is extracted from the biosphere, and 
final energy is sold to consumers.)
But consumers don't buy final energy for its own sake.
They buy final energy so they can convert it to useful energy 
(heat, light, work) that, 
when combined with infrastrucure 
(buildings, roads, etc.),
provides energy services
(heated and cooled space, passenger and freight transport, etc.).

So some energy conversion chain analyses extend the final energy provided by the IEA 
to useful energy through a process that involves

(a) allocating final energy to final-to-useful energy conversion machines,
(b) defining the efficiency of those final-to-useful energy conversion machines, and
(c) calculating the useful energy injected into the economy.

The `IEATools` package provides several functions to assist with the 
final-to-useful analysis process, and
this vignette demonstrates their use.


## Allocation

Extending the IEA's extended energy balance data to the useful stage begins 
with defining allocations from final energy carriers to final-to-useful energy conversion machines.
To demonstrate this process, we'll use the data bundled with this package.

```{r}
library(dplyr)
library(IEATools)
Tidy_iea_data <- load_tidy_iea_df() %>% 
  specify_all() %>% 
  glimpse()
```

The next step is to develop a blank final-to-useful allocation table.

```{r}

```