---
title: "Fixes for IEA Data"
author: "Matthew Kuperus Heun"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Fixes for IEA Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(dplyr)
library(IEATools)
```

## Introduction

Some data in the IEA extended energy balances is incomplete or incorrect.
The `IEATools` package provides functions to fix the incomplete or incorrect data.
These fixes emerge from detailed country studies that lead to academic papers.


## The fixes

At present, there are two fixes to IEA data.


### Ghana Primary solid biofuels

Ghana's Primary solid biofuels data show a very large and dramatic decline from 1999 to 2000.
This decline is due to new survey data being used for the 2000 data.  
When we look at the PSB data on a per-capita basis, it is clear that 
a near-constant PSB/capita value was used to extrapolate per-capita usage in the late 1990s.
When new survey data became available for the 2000 reporting year, 
the per-capita consumption of PSB obviously changed.  
Our approach to this problem is to smooth out the large peak in PSB consumption 
by reducing the per-capita consumption of PSB
in the years 1991--1999.
See the Supplementary material (especially section S2.7) for   

M. K. Heun and P. E. Brockway. 
Meeting 2030 primary energy and economic growth goals: Mission impossible? 
Applied Energy, 251(112697):1–24, May 2019

available [here](http://www.doi.org/10.1016/j.apenergy.2019.01.255)
for additional details.

An example data frame can be constructed 
from the example data supplied with this package
using `load_tidy_iea_df()`.
We pretend that 1971 is 1991 and 2000 is 1992.

```{r}
example_tidy_iea_df <- load_tidy_iea_df() %>% 
  dplyr::filter(Country == "GHA") %>% 
  dplyr::filter(Product == "Primary solid biofuels") %>% 
  dplyr::mutate(
    Year := dplyr::case_when(
      Year == 1971 ~ 1991,
      Year == 2000 ~ 1992
    )
  ) 
example_tidy_iea_df %>% 
  dplyr::filter(Product == "Primary solid biofuels") %>% 
  dplyr::select(Year, Flow.aggregation.point, Flow, E.dot, Unit)
fixed <- example_tidy_iea_df %>% 
  fix_GHA_psb()
```

Comparing production of Primary solid biofuels in 1991
shows that production rates have changed.

```{r}
example_tidy_iea_df %>% 
  dplyr::filter(Year == 1991, Flow == "Production") %>% 
  dplyr::select("E.dot", "Unit")
fixed %>% 
  dplyr::filter(Year == 1991, Flow == "Production") %>% 
  dplyr::select("E.dot", "Unit")
```

The private object `Fixed_GHA_PSB` contains the replacement Primary solid biofuels data.


### Ghana Industry Electricity

IEA data for Ghana begins in 1971.
For 1971--1973, the IEA maintained specificity for Electricity consumption in Industry, including

* Mining and quarrying,
* Non-ferrous metals, and 
* Textile and leather.

All other Electricity consumed in Industry is classified as "Industry not specified elsewhere".

However, annual reports from the Ghana Grid Company ([GridCo](http://www.gridcogh.com)) and 
the Volta River Authority ([VRA](https://www.vra.com)) provide a means to continue providing 
specifics from 1974 to today.
Gaining this specificity is especially helpful for the "Non-ferrous metals" category,
because the pot lines at the Volta Aluminum Company ([VALCO](http://www.valcotema.com))
convert electricity into high-temperature heat for aluminum smelting operations, 
injecting large amounts of high-exergy heat into the economy.

To add specificity to Ghana's Industry Electricity consumption, 
use the `fix_GHA_industry_electricity()` function
with the example data supplied with the package
(`load_tidy_iea_df()`).

```{r}
example_tidy_iea_df <- load_tidy_iea_df() %>% 
  dplyr::filter(Country == "GHA")
fixed <- example_tidy_iea_df %>% 
  fix_GHA_industry_electricity()
```

The 1973 data already have specific Industry Electricity consumption.,
but the 2000 data do not.

```{r}
example_tidy_iea_df %>% 
  dplyr::filter(Flow %in% c("Mining and quarrying",
                     "Non-ferrous metals",
                     "Textile and leather", 
                     "Industry not elsewhere specified"), 
         Product == "Electricity") %>% 
  dplyr::select("Year", "Flow", "E.dot", "Unit")
```

The "fixed" data have added specificity.

```{r}
fixed %>% 
  dplyr::filter(Flow %in% c("Mining and quarrying",
                     "Non-ferrous metals",
                     "Textile and leather", 
                     "Industry not elsewhere specified"), 
         Product == "Electricity") %>% 
  dplyr::select("Year", "Flow", "E.dot", "Unit")
```

Note that the 2000 data are still in balance,
with the sum of 
"Mining and quarrying",
"Non-ferrous metals",
"Textile and leather", and 
"Industry not elsewhere specified" 
*after* fixing 
equal to 
"Industry not elsewhere specified" 
*before* the fixing.

```{r}
example_tidy_iea_df %>% 
  dplyr::filter(Year == 2000, 
         Product == "Electricity", 
         Flow == "Industry not elsewhere specified") %>% 
  dplyr::select(E.dot) %>% 
  as.numeric()
fixed %>% 
  dplyr::filter(Year == 2000, 
         Product == "Electricity", 
         Flow %in% c("Mining and quarrying",
                     "Non-ferrous metals",
                     "Textile and leather", 
                     "Industry not elsewhere specified")) %>% 
  dplyr::select(E.dot) %>% 
  sum()
```

The private object `Fixed_GHA_Industry_Electricity` contains the replacement Industry electricity data.
