---
title: "Prepare for PSUT analysis"
author: "Matthew Kuperus Heun"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prepare for PSUT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
library(IEATools)
library(tidyr)
```

## Introduction

After loading IEA data and ensuring that the data are balanced 
(see the [vignette about loading data](http://IEATools.html)) and
specifying many of the flows
(see the [vignette about specifying flows](http://specify.html)),
the analyst can prepare for physical supply-use table (PSUT) analysis
of an energy conversion chain.
This vignette shows how to use the functions in the `IEATools` package
for that purpose.


## Adding PSUT matrix names (`add_psut_matnames()`)

The first step to converting the IEA extended energy balance data into PSUT matrices
is identifying the matrix in which each datum belongs.
The `add_psut_matnames()` function performs this task.
It adds a column named "matnames" (by default but changeable by the user)
that contains one of `R`, `U`, `V`, or `Y` to identify the 
Resources, Use, Make, or Final demand matrix.

```{r}
load_tidy_iea_df() %>% 
  add_psut_matnames() %>%
  glimpse()
```


## Add matrix metadata (`add_row_col_meta()`)

The next step is to add information to identify the rows and columns 
for each datum. 
The `add_row_col_meta()` function performs this task
by adding `rowname`, `colname`, `rowtype`, and `coltype` 
columns which identify the names and types of rows and columns in the PSUT matrices.
`add_row_col_meta()` should be called after 
calling `add_psut_matnames()`, 
because it relies on the the `matnames` column to 
determine names and types.

```{r}
load_tidy_iea_df() %>% 
  add_psut_matnames() %>%
  add_row_col_meta() %>% 
  glimpse()
```


## Collapse to matrices (`collapse_to_tidy_psut()`)

After identifying row and column names and types, 
the next step is to collapse the data into PSUT matrices.
The `collapse_to_tidy_psut()` function performs this task
using the `matsindf::collapse_to_matrices()` function internally.
The `matvals` column is added to the data frame and
contains a matrix for each set of grouping variables
which are specified in the `grouping_vars` argument.
`collapse_to_tidy_psut()` ensures that all 
entries in each matrix are non-negative.

```{r}
Tidy_PSUT <- 
load_tidy_iea_df() %>% 
  specify_all() %>% 
  add_psut_matnames() %>% 
  add_row_col_meta() %>% 
  collapse_to_tidy_psut()
Tidy_PSUT %>% 
  glimpse()
class(Tidy_PSUT$matvals[[1]])
Tidy_PSUT$matvals[[1]]
```


## Generate `S_units` matrices (`extract_S_units_from_tidy()`)

A matrix of Products and Units can be obtained from a tidy IEA data frame
with the `extract_S_units_from_tidy()` function.

```{r}
S_units_df <- load_tidy_iea_df() %>% 
  extract_S_units_from_tidy()
S_units_df %>% 
  glimpse()
S_units_df$S_units[[1]]
```


## Putting it all together (`prep_psut()`)

`prep_psut()` is a convenience function.
It converts a tidy IEA data frame
that has gone through `specify_all()`
into a PSUT data frame
by collapsing the IEA data into PSUT matrices (R, U, V, and Y).

This function bundles several others:
1. `add_psut_matnames()`
2. `add_row_col_meta()`
3. `collapse_to_tidy_psut()`

Furthermore, it extracts `S_units` matrices using `extract_S_units_from_tidy()`
and adds those matrices to the data frame.

```{r}
Simple <- load_tidy_iea_df() %>% 
  specify_all() %>% 
  prep_psut()
S_units <- load_tidy_iea_df() %>% 
  extract_S_units_from_tidy()
Complicated <- load_tidy_iea_df() %>% 
  specify_all() %>% 
  add_psut_matnames() %>% 
  add_row_col_meta() %>% 
  collapse_to_tidy_psut() %>% 
  spread(key = matnames, value = matvals) %>% 
  full_join(S_units, by = c("Method", "Energy.type", "Last.stage", 
                            "Country", "Year")) %>% 
  gather(key = matnames, value = matvals, R, U_EIOU, U_excl_EIOU, 
         V, Y, S_units)
# Simple and Complicated are same.
full_join(Simple %>% rename(matval_simple = matvals), 
          Complicated %>% rename(matval_complicated = matvals), 
          by = c("Method", "Energy.type", "Last.stage", "Country", "Year", "matnames")) %>% 
  dplyr::mutate(
    same = matsbyname::equal_byname(matval_simple, matval_complicated)
  ) %>% 
  magrittr::extract2("same") %>% 
  as.logical() %>% 
  all()
Simple %>% 
  glimpse()
# U_EIOU for Ghana in 1971
Simple$matvals[[5]]
```


## Conclusion

Taken together, 
the functions in this package make it possible to prepare IEA extended energy balance data
for use with PSUT analysis techniques. 
A sequence of three function calls are all that is required.

```{r}
Tidy_PSUT_mats <- load_tidy_iea_df() %>% 
  specify_all() %>% 
  prep_psut()
Tidy_PSUT_mats %>% 
  glimpse()
```

If calculations are to be performed, the matrices can be spread so that each matrix is in its own column.

```{r}
Tidy_PSUT_mats %>% 
  spread(key = matnames, value = matvals) %>% 
  glimpse()
```
