---
title: "PSUT Names"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PSUT Names}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(IEATools)
```

```{r functions, include = FALSE}
colorize <- function(color, x) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color,
      x)
  } else x
}
```

## Introduction

The `prep_psut()` function creates `matsindf`-style data frames of matrices.
Each matrix has character strings for row and column names.
The row and column names follow strict conventions.
This vignette documents those conventions.


## Rules

The rules for the row and column names are given below:

1. All row and column names consists of a base part (the noun)
   and metadata (usually in the form of prepositional phrases) 
2. The base (noun) always matches the type of thing on that margin.
   E.g., if "Stock changes [of Crude oil]" is a row name in the make (**V**) matrix,
   the base part ("Stock changes") is an industry, as all row names in **V** denote.
3. The metadata are separated from the base by the strings "[" and "]".
3. The metadata can include any number of prepositional phrases.
   In the example above, "of Crude oil" is metadata that modifies
   the base name ("Stock changes")
4. Each prepositional phrase consists of a single-word preposition and 
   its (possibly multi-word) object.
5. Prepositions can be one of "of", "in", "into", "from", "to", and "->". 
   (Note that "->" indicates the outflowing product, 
   a concept for which there is no good preposition in the English language.)


## Matrices

### **R** matrix


#### Rows

The **R** matrix has industries (resources) in rows. 
All row names in the **R** matrix have the same form:

1. Resources: "Resources [of `primary_product`]", 
   where `primary_product` is a primary energy carrier.
   Example: "Resources [of Crude oil]".
   `r colorize("blue", 'With country: "Resources [of Crude oil in SAU]".')`


#### Columns

The **R** matrix has products (energy carriers) in columns.
All column names in the **R** matrix have the same form:

1. Primary products: "`primary_product` [from Resources]",
   where `primary_product` is a primary energy carrier.
   Example: "Crude oil [from Resources]".
   `r colorize("blue", 'With country: "Crude oil [from Resources in SAU]".')`


### **V** matrix


#### Rows

Like the **R** matrix, the **V** matrix has
industries (processing stages) in rows.
There are several options for names of rows in the **V** matrix.

1. Imports: "Imports [of `product`]", where `product` is a primary or final
   energy product.
   Example: "Imports [of Crude oil]".
  `r colorize("blue", 'With country: "Imports [of Crude oil into USA]".')`
2. Stock changes: "Stock changes [of `product`]", where `product` is a primary or 
   final energy carrier.
   Example: "Stock changes [of Crude oil]".
  `r colorize("blue", 'With country: "Stock changes [of Crude oil in SAU]".')`
3. Transformation processes: "`pf_machine`", where `pf_machine` is 
   a machine that converts primary energy to final energy. 
   These data come from the IEA's extended energy balances.
   Example: "Oil refineries".
  `r colorize("blue", 'With country: "Oil refineries [in SAU]".')`
4. Statistical differences: "Statistical differences" which originate from the IEA's
   extended energy balances.
   Example: "Statistical differences". 
   `r colorize("blue", 'With country: "Statistical differences [in SAU]".')`
5. Manufacture processes: "Manufacture [of `product`]", where `product` is any 
   energy carrier. These processes are inserted to convert
   Resources into primary products.
   Example: "Manufacture [of Hydro]".
  `r colorize("blue", 'With country: "Manufacture [of Hydro in ISL]".')`
6. Final-to-useful machines: "`fu_machine` -> `useful_product`", 
   where `fu_machine` is the name of a machine that converts final energy 
   into useful energy and `useful_product` is the name of a useful energy product.
   Example: "LPG stoves -> MTH.200.C".
   `r colorize("red", 'fu_machine [-> useful_product]')`
  `r colorize("blue", 'With country: "LPG stoves [-> MTH.200.C in GHA]".')`


#### Columns

Like the **R** matrix, the **V** matrix has
products (energy carriers) in columns. 
There are two options for names of columns in the **V** matrix.

1. Primary and final energy carriers: "`product`" where `product` is the 
   unmodified name of the energy carrier in the IEA extended energy balances.
   Example: "Anthracite".
  `r colorize("blue", 'With country: "Anthracite [in AUS]".')`
2. Useful products: "`useful_product` [from `fu_machine`]", 
   where `useful_product` is a useful energy carrier and `fu_machine` is a 
   final-to-useful machine.
   Example: "HPA [from Air compressors]"
   `r colorize("blue", 'With country: "HPA [from Air compressors in USA]".')`



### **U** matrix


#### Rows

The **U** matrix has products (energy carriers) in rows.

1. Resource flows: "`primary_product` [from Resources]", where
   `primary_product` is a primary energy carrier.
   Example: "Biogasoline [from Resources]".
  `r colorize("blue", 'With country: "Biogasoline [from Resources in USA]".')`
2. Primary and final energy flows: "`product`", where product is a primary
   or final energy carrier.
   Example: "Electricity".
  `r colorize("blue", 'With country: "Electricity [in GBR]".')`
3. Useful products: "`useful_product` [from `fu_machine`]"], where
   `useful_product` is a useful energy carrier and 
   `fu_machine` is a final-to-useful machine.
   Example: "HPL [from Electric pumps]".
  `r colorize("blue", 'With country: "HPL [from Electric pumps in ESP]".')`


#### Columns

The **U** matrix has industries (processing stages) in columns.

1. Primary-to-final machines: "`pf_machine`" where `pf_machine` is the name
   of a primary-to-final machine.
   Example: "Blast furnaces"
  `r colorize("blue", 'With country: "Blast furnaces [in USA]".')`
2. Final-to-useful machines: "`fu_machine` -> `useful_product`", where
   `fu_machine` is a final-to-useful machine and 
   `useful_product` is a useful energy carrier. 
   `r colorize("red", 'fu_machine [-> useful_product]')`
   `r colorize("blue", 'With country: "Ammonia synthesizers [-> HTH.400.C in POL]".')`


### **Y** matrix

#### Rows

Like the **U** matrix, the **Y** matrix has products (energy carriers) in rows.

1. Primary and final products: "`product`", where `product` is the 
   unmodified name of the energy carrier in the IEA extended energy balances.
   Example: "Naptha".
  `r colorize("blue", 'With country: "Naptha [in ESP]".')`
2. Useful products: "`product` [from `fu_machine`]", where `product` is a useful
   energy carrier and `fu_machine` is a final-to-useful machine.
   Example: "RoP [from Electric cars]".
  `r colorize("blue", 'With country: "RoP [from Electric cars in USA]".')`
  

#### Columns

Like the **U** matrix, the **Y** matrix has industries (final demand sectors) in columns.

1. Sectors: "`sector`", where `sector` is the name of a final demand sector, 
   unmodified from the IEA extended energy balances. 
   Example: "Construction".
  `r colorize("blue", 'With country: "Construction [in USA]".')`
2. Exports: "Exports [of `product`]", where `product` is a primary or final
   energy product.
   Example: "Exports [of Coking coal]".
  `r colorize("blue", 'With country: "Exports [of Coking coal from USA]".')`
3. Exports to World marine/aviation bunkers: 
   "Exports to World `bunker_type` bunkers [of `product`]", where
   `bunker_type` is either "marine" or "aviation" and `product` is a 
   (typically) final energy product.
   Example: "Exports to World marine bunkers [of Biodiesels]".
  `r colorize("blue", 'With country: "Exports to World marine bunkers 
  [of Biodiesels from USA]".')`
   

 