---
title: "Specify EIOU Industries and Flows"
author: "Matthew Kuperus Heun"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Specify EIOU Industries and Flows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
library(IEATools)
library(magrittr)
```

## Introduction

The IEA extended energy balance data include 
some `Flow`s that are `Energy industry own use` (EIOU), 
the consumption of energy by energy-producing industries.
This vignette demonstrates how to deal with the EIOU flows,
especially if an analyst is moving toward performing 
Physical Supply-Use Table (PSUT) analyses
with the IEA data.
In particular, this vignette will identify some of the problems with using such flows
(as they are present in data from the IEA)
and how to use functions in the `IEATools` package to 
prepare for PSUT analysis. 

We'll use data supplied with this package and 
accessed by the `load_tidy_iea_df()` function
to illustrate.


### The IEA approach to Energy Industry Own Use (EIOU) flows

The EIOU flows can be found as shown in the following code.
The `Flow` column indicates the industry to which EIOU of type `Product` is flowing.

```{r}
library(dplyr)
library(IEATools)
library(magrittr)
EIOU_flows <- load_tidy_iea_df() %>% 
  filter(Flow.aggregation.point == "Energy industry own use")
EIOU_flows %>% 
  select(Country, Year, Flow, Product)
```

And `EIOU`-consuming industries can be found by the following code.

```{r}
EIOU_flows$Flow %>% unique()
```

Note that one of the industries that receives EIOU is `Coal mines (energy)`, 
but coal is not produced by `Coal mines (energy)` in the IEA data. 
(The suffix `(energy)` indicates EIOU.)
Rather, coal (of various types) first appears in 
rows where the `Flow` is `Production`.

```{r}
load_tidy_iea_df() %>% 
  filter(Flow.aggregation.point == "Total primary energy supply", 
         Product %in% coal_and_coal_products) %>% 
  select(Country, Flow.aggregation.point, Flow, Product) %>% 
  unique()
```


### Problems with the IEA approach to EIOU

There are two problems with the IEA's approach to energy industry own use (EIOU).

* Assigning all originating flows to `Production` (as the IEA does)
  means that some EIOU is routed to an industry that does not produce anything.
  In the example above, `Electricity` is routed to `Coal mines (energy)`, but
  coal is produced by `Production`.
* Furthermore, if all primary energy carriers originate at `Production`,
  there will be problems with "upstream swims" in an Input-Output analysis:
  demand for one primary energy carrier will imply proportional demand 
  for all other primary energy carriers.


### Solving the problems presented by the IEA's approach to EIOU

To solve these problems,
generic `Production` `Flow`s need to be specified.
In the example above, 
`coal_and_coal_product`s should be produced by `Coal mines`, 
not the generic `Production` industry. 
The `specify_primary_production()` function accomplishes this task.

## Identify primary production (`specify_primary_production()`)

By default, the `specify_primary_produciton()` function
performs the following actions:

* Coal and coal products
    * The `Production` industry for `coal_and_coal_products` 
      is replaced by `Coal mines`.
    * A `Resources (Coal)` industry is created which becomes the 
      source of all primary `coal_and_coal_products` in the energy conversion chain.
    * Flows from `Resources (Coal)` to `Coal mines` are added, 
      such that `Resources (Coal)` makes primary `coal_and_coal_products` but 
      doesn't use any energy.
    * The `Coal mines (energy)` EIOU `Flow` is replaced by `Coal mines`.
* Oil and oil products and Natural gas
    * The `Production` industry for `oil_and_oil_products` and `Natural gas`
      is replaced by `Oil and gas extraction`.
    * A `Resources (Oil and natural gas)` industry is created which becomes the 
      source of all primary `oil_and_oil_products` and `Natural gas` in the energy conversion chain.
    * Flows from `Resources (Oil and natural gas)` to `Oil and gas extraction` are added,
      such that `Resources (Oil and natural gas)` makes primary `Oil_and_oil_products` and `Natural gas`
      but doesn't use any energy.
    * The `Oil and gas extraction (energy)` EIOU `Flow` is replaced by `Oil and gas extraction`.

The code below demonstrates `specify_primary_production()`.
Note that the `Resources` `Flow`s are tagged as belonging to the Resources matrix (`R`)
by the `add_psut_matnames()` function.

```{r}
Specific_primary_production <- load_tidy_iea_df() %>% 
  specify_primary_production() %>% 
  add_psut_matnames()
Specific_primary_production %>% 
  filter(Flow %in% c("Resources (Coal)", "Resources (Oil and natural gas)")) %>%
  select(-Method, -Last.stage, -Energy.type, -Ledger.side, -Unit) %>% 
  as.data.frame()
```

Furthermore, the added flows from the `Resource` industries are
tagged as `Transformation processes` and 
belong in the `U_excl_EIOU` and `V` matrices. 
The EIOU flows are shown to belong in the `U_EIOU` matrix.

```{r}
Specific_primary_production %>% 
  filter(Flow %in% c("Coal mines", "Oil and gas extraction")) %>% 
  as.data.frame()
Specific_primary_production %>% 
  filter(Flow.aggregation.point == "Energy industry own use" ) %>% 
  select(Country, Year, Flow, Product, E.dot, matnames) %>% 
  as.data.frame()
```


## Convert `Production` to `Resources` (`specify_production_to_resources()`)

The IEA extended energy balances give resource extraction in rows where the `Flow` is "`Production`".
`specify_production_to_resources()` changes the "`Production`" string 
to "`Resources (product)`", 
where `product` is the name of the energy carrier for this resource.
`specify_production_to_resources()` should be called _after_ `specify_primary_production()`,
which adjusts for energy industry own use 
of some primary energy producing industries.
If this function is called first, 
energy industry own use will not be accounted correctly.

```{r}
load_tidy_iea_df() %>% 
  specify_primary_production() %>% 
  specify_production_to_resources() %>% 
  filter(startsWith(Flow, "Resources")) %>% 
  glimpse()
```


## Be more specific about interface industries (`specify_interface_industries()`)

An interface industry is one that moves energy carriers into or out of a country.
When `Flow` is any of the interface industries, we need to be more specific.
If we don't separate these `Flow`s by `Product`, we run into trouble 
in the PSUT framework:

* upstream swims will result in all `Product`s being produced even if only one is needed and
* embodied energy calculations will result in many types of energy being embodied, even if only one sould be.

To solve these problems, `specify_interface_industries()`
adds a suffix ` (Product)` to each of these interface industries.
 
Note that "`Production`" `Flow`s also needs to be specified, 
but that is accomplished in the `specify_primary_production()` and
`specify_production_to_resources()` functions.

```{r}
load_tidy_iea_df() %>% 
  specify_interface_industries() %>% 
  filter(starts_with_any_of(Flow, IEATools::interface_industries)) %>% 
  glimpse()
```


## Be more specific about Transformation process Energy industry own use (`specify_tp_eiou()`)

The extended energy balance data from the IEA includes 
"Energy industry own use" (EIOU) for many transformation processes.
Unfortunately, in some cases
the EIOU flows into industries that aren't included in transformation processes.
For example, `Electricity` is consumed by 
`Own use in electricity, CHP and heat plants`, 
which is not a transformation process.
We have to make some decisions to ensure that 
EIOU is routed to actual transformation processes.
Changes made to the `.tidy_iea_df` data frame are:
 
1. EIOU classified as `own_use_elect_chp_heat` (by default "Own use in electricity, CHP and heat plants")
   is sent to `main_act_producer_elect` (by default "Main activity producer electricity plants").
2. EIOU classified as `pumped_storage` (by default "Pumped storage plants"") 
   is sent to `main_act_producer_elect`.
3. EIOU classified as `nuclear_industry` (by default "Nuclear industry") is sent to `main_act_producer_elect`.
4. EIOU classified as `non_spec_energy` (by default "Non-specified (energy)") 
   is sent to `nonspecenergy_reclassify` (by default "Non-specified (transformation)").

```{r}
load_tidy_iea_df() %>% 
  specify_tp_eiou() %>% 
  filter(Flow.aggregation.point == "Energy industry own use" & 
           Flow == "Main activity producer electricity plants") %>% 
  glimpse()
```


## Identify Transformation process sinks and sources (`tp_sinks_sources()`)

In the IEA extended energy balance data, 
transformation processes (tp) ought to both consume and produce energy.
But some transformation processes consume energy without producing any energy; 
others produce without consuming.
Such transformation processes can be called "transformation sinks" and 
"transformation sources," respectively.
This function finds and identifies transformation processes that act as sinks or sources.

It is important to identify transformation sinks, 
because they cause two problems for physical supply-use table (PSUT) analysis. 
First, when swimming upstream, a PSUT analysis cannot "see" the sunk energy carriers,
because they have no downstream effects.
Thus, upstream swims cannot conserve energy.
Second, when calculating embodied energy for each downstream energy carrier,
the sunk energy carriers cannot be embodied in any final demand energy carriers.
Thus, embodied energy calculations cannot conserve energy.
 
Transformation sources can also cause problems for physical supply-use table (PSUT) analysis. 
In particular, when swimming upstream, a PSUT analysis will "see" the final energy sources,
but cannot see any associated primary energy carriers.

Transformation sinks and sources are identified by the following algorithm:
 
1. Identify (per group in `.tidy_iea_df`) all `Transformation processes` that consume energy (negative value for `E.dot`).
   Energy consumption can be for the transformation process itself or for Energy industry own use.
2. Identify (per group in `.tidy_iea_df`) all `Transformation processes` that produce energy (positive value for `E.dot`).
3. Take the set difference between the two (consumers less producers for sinks and producers less consumers for sources). 
   The set difference is the list of transformation sinks or sources, respectively.
 
`tp_sinks_sources()` is a function not unlike `dplyr::summarise()`;
it returns a summary containing grouping variables and industries that are transformation sinks or sources.
The various `specify_*()` functions should also be called _before_ calling [tp_sinks_sources()].
The `specify_*()` functions clean up the IEA data, ensuring that energy is routed to the right places.
 
Note that this function only identifies transformation sinks or sources;
it does not fix the problem. 
To solve the problem of transformation sinks, 
see the [tp_sinks_to_nonenergy()] function.
`tp_sinks_to_nonenergy()` uses the output of `tp_sinks_sources()`
to route energy consumed by transformation sinks to `Non-energy use industry/transformation/energy`.
There is no function to solve the problem of transformation sources at this time.

`tp_sinks_sources()` returns metadata columns and the `Flow` column.
The `Flow` column contains industries that are sinks or sources,
depending on the value of the `type` argument.

```{r}
# The included data sets to not have any transformation process sinks or sources.
# Demonstrate with a made-up data set.
data.frame(Flow = c("Automobiles", "Automobiles", "Furnaces"),
                   E.dot = c(-1, 1, 2), 
                   stringsAsFactors = FALSE) %>% 
  dplyr::mutate(
    Country = "A country",
    Flow.aggregation.point = "Transformation processes",
    Product = "A product"
  ) %>% 
  # Automobiles both consume and produce energy, 
  # so Automobiles are not reported by tp_sinks_sources.
  # However, Furnaces make Product without consuming any energy, 
  # and are, therefore, a transformation source.
  tp_sinks_sources(type = "sources")
```


## Convert Transformation process sinks to Non-energy flows (`tp_sinks_to_nonenergy()` )

This function reclassifies energy flowing into transformation process sinks
as `non_energy_flow`, by default "Non-energy use in industry/transformation/energy".
If there was already some Non-energy use, 
the new Non-energy use is added to the existing Non-energy use.

`tp_sinks_to_nonenergy()` uses the `tp_sinks_sources()` function
internally to identify transformation process sinks.

```{r}
DF <- data.frame(
  Ledger.side = c("Supply", "Supply", "Supply", "Consumption"),
  Flow.aggregation.point = c("Transformation processes", 
                             "Transformation processes", 
                             "Transformation processes", 
                             "Non-energy use"), 
  Flow = c("Automobiles", "Automobiles", "Furnaces", 
           "Non-energy use industry/transformation/energy"),
  Product = c("Petrol", "MD", "Coal", "Coal"),
  E.dot = c(-1, 1, -2, 8), 
  stringsAsFactors = FALSE
) %>% 
  mutate(
    Method = "PCM", 
    Last.stage = "Final",
    Energy.type = "E",
    Country = "Bogus",
    Year = 1971
  )
DF
# In this example, Furnaces are a transformation process sink:
# they consume Coal but produce nothing.
# There are already 8 units of Coal consumption for Non-energy uses.
# The 2 units of Coal are added to the existing 8 units
# Non-energy consumption to make 10 units of Non-energy consumption. 
DF %>% 
  tp_sinks_to_nonenergy() %>% 
  as.data.frame()
```


## Put it all together (`specify_all()`)

`specify_all()` is a convenience function that 
bundles several others:

1. `specify_primary_production()`
2. `specify_production_to_resources()`
3. `specify_tp_eiou()`
4. `specify_interface_industries()`
5. `tp_sinks_to_nonenergy()`

Each bundled function is called in turn using default arguments.

```{r}
Simple <- load_tidy_iea_df() %>% 
  specify_all()
Complicated <- load_tidy_iea_df() %>% 
  specify_primary_production() %>% 
  specify_production_to_resources() %>% 
  specify_tp_eiou() %>% 
  specify_interface_industries() %>% 
  tp_sinks_to_nonenergy()
all(Simple == Complicated)
```

## Conclusion

This vignette demonstrated how to specify details of primary and transformation process flows.
The next step could be to prepare the IEA data for use within the PSUT framework.
For details, see the [Prep PSUT](prep_psut.html) vignette.
