---
title: "Specify EIOU Industries and Flows"
author: "Matthew Kuperus Heun"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Specify EIOU Industries and Flows}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
library(IEATools)
library(magrittr)
```

## Introduction

The IEA extended energy balance data include some `Flow`s that are
`Energy industry own use` (EIOU), the consumption of energy by
energy-producing industries. This vignette demonstrates how to deal with
the EIOU flows, especially if an analyst is moving toward performing
Physical Supply-Use Table (PSUT) analyses with the IEA data. In
particular, this vignette will identify some of the problems with using
such flows (as they are present in data from the IEA) and how to use
functions in the `IEATools` package to prepare for PSUT analysis.

We'll use data supplied with this package and accessed by the
`load_tidy_iea_df()` function to illustrate.

### The IEA approach to Energy Industry Own Use (EIOU) flows

The EIOU flows can be found as shown in the following code. The `Flow`
column indicates the industry to which EIOU of type `Product` is
flowing.

```{r}
library(dplyr)
library(IEATools)
library(magrittr)
EIOU_flows <- load_tidy_iea_df() %>%
  filter(Flow.aggregation.point == "Energy industry own use")
EIOU_flows %>%
  select(Country, Year, Flow, Product)
```

And `EIOU`-consuming industries can be found by the following code.

```{r}
EIOU_flows$Flow %>% unique()
```

Note that one of the industries that receives EIOU is
`Coal mines (energy)`, but coal is not produced by `Coal mines (energy)`
in the IEA data. (The suffix `(energy)` indicates EIOU.) Rather, coal
(of various types) first appears in rows where the `Flow` is
`Production`.

```{r}
load_tidy_iea_df() %>%
  filter(Flow.aggregation.point == "Total primary energy supply",
         Product %in% coal_and_coal_products) %>%
  select(Country, Flow.aggregation.point, Flow, Product) %>%
  unique()
```

### Problems with the IEA approach to EIOU

There are two problems with the IEA's approach to energy industry own
use (EIOU).

-   Assigning all originating flows to `Production` (as the IEA does)
    means that some EIOU is routed to an industry that does not produce
    anything. In the example above, `Electricity` is routed to
    `Coal mines (energy)`, but coal is produced by `Production`.
-   Furthermore, if all primary energy carriers originate at
    `Production`, there will be problems with "upstream swims" in an
    Input-Output analysis: demand for one primary energy carrier will
    imply proportional demand for all other primary energy carriers.

### Solving the problems presented by the IEA's approach to EIOU

To solve these problems, the following fixed need to be implemented:

-   First, generic `Production` `Flow`s need to be specified. In the
    example above, `coal_and_coal_product`s should be produced by
    `Coal mines`, not the generic `Production` industry. Likewise,
    `oil_and_gas_product`s should be produced by
    `Oil and gas extraction`. For all other products, the production
    industry is named as `Manufacture \[of product\]`.
-   Second, the resources matrix formulation needs to be implemented,
    meaning that `Production` `Flow`s need to come from resources
    industries, which supply resource-products, that are then used in
    each of the manufacturing industries (`Coal mines`,
    `Oil and gas extraction`, and `Manufacture \[of product\]`) for
    manufacturing the actual products.

The `specify_primary_production()` function accomplishes these tasks.

## Specify primary production (`specify_primary_production()`)

By default, the `specify_primary_produciton()` function performs the
following actions:

-   First, `Production` `Flow`s for a given `Product` are duplicated.

-   Second, the first duplicate of `Production` `Flow`s for a given
    `Product` are replaced by a `Resource \[of Product\]` `Flow`, which
    produces `Product \[from Resources\]`.

-   Third, the second duplicate of `Production` `Flow`s for a given
    `Product` are replaced by a manufacturing industry flow, that takes
    as input the `Product \[from Resources\]` supplied by the new
    `Resource \[of Product\]` `Flow`, and that supplies as output the
    actual `Product`. The manufacturing industry is named:

    -   `Coal mines` in the case of `coal_and_coal_products`;
    -   `Oil and gas extraction` in the case of `oil_and_oil_products`;
    -   `Manufacture \[of Product\]` for all other products.

-   

-   Coal and coal products

    -   The `Production` industry for `coal_and_coal_products` is
        replaced by `Coal mines`.
    -   A `Resources (Coal)` industry is created which becomes the
        source of all primary `coal_and_coal_products` in the energy
        conversion chain.
    -   Flows from `Resources (Coal)` to `Coal mines` are added, such
        that `Resources (Coal)` makes primary `coal_and_coal_products`
        but doesn't use any energy.
    -   The `Coal mines (energy)` EIOU `Flow` is replaced by
        `Coal mines`.

-   Oil and oil products and Natural gas

    -   The `Production` industry for `oil_and_oil_products` and
        `Natural gas` is replaced by `Oil and gas extraction`.
    -   A `Resources (Oil and natural gas)` industry is created which
        becomes the source of all primary `oil_and_oil_products` and
        `Natural gas` in the energy conversion chain.
    -   Flows from `Resources (Oil and natural gas)` to
        `Oil and gas extraction` are added, such that
        `Resources (Oil and natural gas)` makes primary
        `Oil_and_oil_products` and `Natural gas` but doesn't use any
        energy.
    -   The `Oil and gas extraction (energy)` EIOU `Flow` is replaced by
        `Oil and gas extraction`.

The code below demonstrates `specify_primary_production()`. Note that
the `Resources` `Flow`s are tagged as belonging to the Resources matrix
(`R`) by the `add_psut_matnames()` function.

```{r}
Specific_primary_production <- load_tidy_iea_df() %>%
  specify_primary_production() %>%
  add_psut_matnames()
Specific_primary_production %>%
  filter(Flow %in% c("Resources (Coal)", "Resources (Oil and natural gas)")) %>%
  select(-Method, -Last.stage, -Energy.type, -Ledger.side, -Unit) %>%
  as.data.frame()
```

Furthermore, the added flows from the `Resource` industries are tagged
as `Transformation processes` and belong in the `U_feed` and `V`
matrices. The EIOU flows are shown to belong in the `U_EIOU` matrix.

```{r}
Specific_primary_production %>%
  filter(Flow %in% c("Coal mines", "Oil and gas extraction")) %>%
  as.data.frame()
Specific_primary_production %>%
  filter(Flow.aggregation.point == "Energy industry own use" ) %>%
  select(Country, Year, Flow, Product, E.dot, matnames) %>%
  as.data.frame()
```

## Convert `Production` to `Resources` (`specify_production_to_resources()`)

The IEA extended energy balances give resource extraction in rows where
the `Flow` is "`Production`". `specify_production_to_resources()`
changes the "`Production`" string to "`Resources (product)`", where
`product` is the name of the energy carrier for this resource.
`specify_production_to_resources()` should be called *after*
`specify_primary_production()`, which adjusts for energy industry own
use of some primary energy producing industries. If this function is
called first, energy industry own use will not be accounted correctly.

```{r}
load_tidy_iea_df() %>%
  specify_primary_production() %>%
  specify_production_to_resources() %>%
  filter(startsWith(Flow, "Resources")) %>%
  glimpse()
```

## Be more specific about interface industries (`specify_interface_industries()`)

An interface industry is one that moves energy carriers into or out of a
country. When `Flow` is any of the interface industries, we need to be
more specific. If we don't separate these `Flow`s by `Product`, we run
into trouble in the PSUT framework:

-   upstream swims will result in all `Product`s being produced even if
    only one is needed and
-   embodied energy calculations will result in many types of energy
    being embodied, even if only one should be.

To solve these problems, `specify_interface_industries()` adds a suffix
`(Product)` to each of these interface industries.

Note that "`Production`" `Flow`s also needs to be specified, but that is
accomplished in the `specify_primary_production()` and
`specify_production_to_resources()` functions.

```{r}
load_tidy_iea_df() %>%
  specify_interface_industries() %>%
  filter(starts_with_any_of(Flow, IEATools::interface_industries)) %>%
  glimpse()
```

## Be more specific about Transformation process Energy industry own use (`specify_tp_eiou()`)

The extended energy balance data from the IEA present some issues when
linking "Energy Industry Own Use" (EIOU) flows to "Transformation
processes" flows:

1.  In some instances the data includes "Energy industry own use" (EIOU)
    flows for industries that are not described as transformation
    processes. For instance, the "Nuclear industry" appears as an EIOU
    flow but does not appear as a transformation process.
2.  In some other instances the data includes transformation processes
    for which no "Energy Industry Own Use" flow is reported. For
    instance, the "Main activity producer electricity plants" is a
    transformation process, but no EIOU is associated.

The `specify_tp_eiou` function enables to deal with these
inconsistencies. Effectively, it is a convenience function that bundles
together several other functions:

1.  `gather_producer_autoproducer()`
2.  `route_pumped_storage()`
3.  `route_own_use_elect_chp_heat()`
4.  `add_nuclear_industry()`
5.  `route_non_specified_eiou()`
6.  `route_non_specified_tp()`

Details can be found in the documentation of each function, but
essentially they do the following:

1.  The `gather_producer_autoproducer()` function gathers *separately*
    the transformation processes "Main activity producer plants" and
    "Autoproducer plants" for each of three types: electricity, heat,
    and CHP.
2.  The `route_pumped_storage()` function routes the "Pumped storage
    plants" EIOU flow to "Main activity producer electricity plants."
3.  The `route_own_use_elect_chp_heat()` routes the "Own use in
    electricity, heat and CHP plants" EIOU flow to "Main activity
    producer electricity plants", "Main activity producer CHP plants,"
    and "Main activity producer heat plants." It does so according to
    either the input or output shares of each of the industries,
    depending on the arguments of the function.
4.  The `add_nuclear_industry()` function adds a nuclear industry in
    transformation processes, and modifies each of the "Main activity
    producer electricity plants" and "Main activity producer CHP plants"
    according to their inputs of nuclear fuel as reported by the IEA.
5.  The `route_non_specified_eiou()` function routes the "Non-specified"
    EIOU flow to the different EIOU industries. It performs the split
    according to the shares of use of EIOU of all EIOU industries
    (excluding the non-specified one).
6.  The `route_non_specified_tp()` function routes the "Non-specified"
    transformation processes flows to the different transformation
    processes industries. It performs the split according to the shares
    of use and supply of each product, by each transformation process
    industry (excluding the non-specified one).

The following two pieces of code are therefore equivalent:

```{r}
load_tidy_iea_df() %>%
  specify_tp_eiou() %>%
  glimpse()

load_tidy_iea_df() %>% 
  gather_producer_autoproducer() %>% 
  route_pumped_storage() %>% 
  route_own_use_elect_chp_heat() %>% 
  add_nuclear_industry() %>% 
  route_non_specified_eiou() %>% 
  route_non_specified_tp() %>% 
  glimpse()
```

## Identify Transformation process sinks and sources (`tp_sinks_sources()`)

In the IEA extended energy balance data, transformation processes (tp)
ought to both consume and produce energy. But some transformation
processes consume energy without producing any energy; others produce
without consuming. Such transformation processes can be called
"transformation sinks" and "transformation sources," respectively. This
function finds and identifies transformation processes that act as sinks
or sources.

It is important to identify transformation sinks, because they cause two
problems for physical supply-use table (PSUT) analysis. First, when
swimming upstream, a PSUT analysis cannot "see" the sunk energy
carriers, because they have no downstream effects. Thus, upstream swims
cannot conserve energy. Second, when calculating embodied energy for
each downstream energy carrier, the sunk energy carriers cannot be
embodied in any final demand energy carriers. Thus, embodied energy
calculations cannot conserve energy.

Transformation sources can also cause problems for physical supply-use
table (PSUT) analysis. In particular, when swimming upstream, a PSUT
analysis will "see" the final energy sources, but cannot see any
associated primary energy carriers.

Transformation sinks and sources are identified by the following
algorithm:

1.  Identify (per group in `.tidy_iea_df`) all
    `Transformation processes` that consume energy (negative value for
    `E.dot`). Energy consumption can be for the transformation process
    itself or for Energy industry own use.
2.  Identify (per group in `.tidy_iea_df`) all
    `Transformation processes` that produce energy (positive value for
    `E.dot`).
3.  Take the set difference between the two (consumers less producers
    for sinks and producers less consumers for sources). The set
    difference is the list of transformation sinks or sources,
    respectively.

`tp_sinks_sources()` is a function not unlike `dplyr::summarise()`; it
returns a summary containing grouping variables and industries that are
transformation sinks or sources. The various `specify_*()` functions
should also be called *before* calling [tp_sinks_sources()]. The
`specify_*()` functions clean up the IEA data, ensuring that energy is
routed to the right places.

Note that this function only identifies transformation sinks or sources;
it does not fix the problem. To solve the problem of transformation
sinks, see the [tp_sinks_to_nonenergy()] function.
`tp_sinks_to_nonenergy()` uses the output of `tp_sinks_sources()` to
route energy consumed by transformation sinks to
`Non-energy use industry/transformation/energy`. There is no function to
solve the problem of transformation sources at this time.

`tp_sinks_sources()` returns metadata columns and the `Flow` column. The
`Flow` column contains industries that are sinks or sources, depending
on the value of the `type` argument.

```{r}
# The included data sets to not have any transformation process sinks or sources.
# Demonstrate with a made-up data set.
data.frame(Flow = c("Automobiles", "Automobiles", "Furnaces"),
                   E.dot = c(-1, 1, 2),
                   stringsAsFactors = FALSE) %>%
  dplyr::mutate(
    Country = "A country",
    Flow.aggregation.point = "Transformation processes",
    Product = "A product"
  ) %>%
  # Automobiles both consume and produce energy,
  # so Automobiles are not reported by tp_sinks_sources.
  # However, Furnaces make Product without consuming any energy,
  # and are, therefore, a transformation source.
  tp_sinks_sources(type = "sources")
```

## Convert Transformation process sinks to Non-energy flows (`tp_sinks_to_nonenergy()` )

This function reclassifies energy flowing into transformation process
sinks as `non_energy_flow`, by default "Non-energy use in
industry/transformation/energy". If there was already some Non-energy
use, the new Non-energy use is added to the existing Non-energy use.

`tp_sinks_to_nonenergy()` uses the `tp_sinks_sources()` function
internally to identify transformation process sinks.

```{r}
DF <- data.frame(
  Ledger.side = c("Supply", "Supply", "Supply", "Consumption"),
  Flow.aggregation.point = c("Transformation processes",
                             "Transformation processes",
                             "Transformation processes",
                             "Non-energy use"),
  Flow = c("Automobiles", "Automobiles", "Furnaces",
           "Non-energy use industry/transformation/energy"),
  Product = c("Petrol", "MD", "Coal", "Coal"),
  E.dot = c(-1, 1, -2, 8),
  stringsAsFactors = FALSE
) %>%
  mutate(
    Method = "PCM",
    Last.stage = "Final",
    Energy.type = "E",
    Country = "Bogus",
    Year = 1971
  )
DF
# In this example, Furnaces are a transformation process sink:
# they consume Coal but produce nothing.
# There are already 8 units of Coal consumption for Non-energy uses.
# The 2 units of Coal are added to the existing 8 units
# Non-energy consumption to make 10 units of Non-energy consumption.
DF %>%
  tp_sinks_to_nonenergy() %>%
  as.data.frame()
```

## Put it all together (`specify_all()`)

`specify_all()` is a convenience function that bundles several others:

1.  `specify_primary_production()`
2.  `specify_production_to_resources()`
3.  `specify_tp_eiou()`
4.  `specify_interface_industries()`
5.  `tp_sinks_to_nonenergy()`

Each bundled function is called in turn using default arguments.

```{r}
Simple <- load_tidy_iea_df() %>%
  specify_all()
Complicated <- load_tidy_iea_df() %>%
  specify_primary_production() %>%
  specify_production_to_resources() %>%
  specify_tp_eiou() %>%
  specify_interface_industries() %>%
  tp_sinks_to_nonenergy()
all(Simple == Complicated)
```

## Conclusion

This vignette demonstrated how to specify details of primary and
transformation process flows. The next step could be to prepare the IEA
data for use within the PSUT framework. For details, see the [Prep
PSUT](https://matthewheun.github.io/IEATools/articles/prep_psut.html)
vignette.
